Follow-up Samples

### 2017-03-08 submission/manuscript_outline_20161024.Rmd
update to include effect sizes of chemo/rads treatment

In this study we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.  We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).  We then investigated several important genera from oral microbes that have been suggested to be important in CRC pathogenesis [@flynn_metabolic_2016].  We next utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up samples.  Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether initial and follow up samples were changing and whether it was towards a normal community.  We also investigated the two models for similar important OTUs to identify the crucial OTUs involved with both classifying lesion or normal and initial versus follow up samples.  This study will provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.

@@

***Carcinoma Associated Bacteria:***
So we next examined whether there were changes in previously well described carcinoma associated OTUs, such as *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  We first observed that only a small percentage of those with adenoma or carcinoma were positive or had a relative abundance above 0.1% for any of these respective OTUs [Figure 2].  Despite this, those with carcinoma had a decrease in relative abundance from initial to follow up for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 2A]. In contrast, there was no difference in relative abundance in any of these OTUs when considering only those with adenoma [Figure 2B].

@@

***Treatment Differences:***
After observing these changes in the bacterial community and positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.  There was only a significant difference for change in positive probability for those treated with chemotherapy for the initial sample model (P-value `r ifelse(chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).  This suggests that follow up samples for those treated for chemotherapy my have had a larger change from the initial sample then those without such treatment.  All other variables that were tested showed no difference based on whether chemotherapy or radiation was used [Table S3].

>>>>

In this study we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.  We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).  We then investigated several important genera from oral microbes that have been suggested to be important in CRC pathogenesis [@flynn_metabolic_2016].  We next utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up samples.  Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether initial and follow up samples were changing and whether it was towards a normal community.  We also investigated the two models for similar important OTUs to identify the crucial OTUs involved with both classifying lesion or normal and initial versus follow up samples.  Finally, since treatment also included chemotherapy and radiation therapy along with lesion removal, we tested if these additionaly treatments significantly lowered any of the metrics examined versus those who received only lesion removal.  This study will provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.

@@

***Carcinoma Associated Bacteria:*** (Anna recommends moving this to the supplement)
So we next examined whether there were changes in previously well described carcinoma associated OTUs, such as *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  We first observed that only a small percentage of those with adenoma or carcinoma were positive or had a relative abundance above 0.1% for any of these respective OTUs [Figure 2].  Despite this, those with carcinoma had a decrease in relative abundance from initial to follow up for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 2A]. In contrast, there was no difference in relative abundance in any of these OTUs when considering only those with adenoma [Figure 2B].

@@

***Treatment Differences:*** (after meeting with Anna an idea I have is that it might be a good idea to tie all the findings together and relate it back to treatment)
After observing these changes in the bacterial community and positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.  There was only a significant difference for change in positive probability for those treated with chemotherapy for the initial sample model (P-value `r ifelse(chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).  This suggests that follow up samples for those treated for chemotherapy my have had a larger change from the initial sample then those without such treatment.  All other variables that were tested showed no difference based on whether chemotherapy or radiation was used [Table S3].
